Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2021-04-12 document/source/abstract.tex
new document build structuring with common source

\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
% \input{archive/main.config.tex}
\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
% \addbibresource{IROS.bib}
\addbibresource{cav_determinism.bib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{balance} %balance columns
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{enumitem}
\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
% \usepackage{subfig}
\usepackage{subcaption}
\usepackage{dblfloatfix}
\usepackage{booktabs}
\usepackage{authblk}
\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
% \pagenumbering{gobble} %suppresses page numbers
\usepackage{balance} %balance columns
%\usepackage{todonotes}
\usepackage[disable]{todonotes}
\usepackage{siunitx}
% \usepackage{draftwatermark}
\usepackage{flushend}

%-------------------------------------------------------------------------------

\begin{document}

% ***************************************************
%  TITLE
% ***************************************************
\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
% \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
% \author[1,3]{Greg Chance}
% \author[1,3]{Kevin McAreavey}
% \author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
% \author[2,3]{Tony Pipe}
% \author[1,3]{Kerstin Eder}
% \affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
% \affil[2]{University of West of England, Bristol, UK}
% \affil[3]{Bristol Robotics Laboratory, Bristol, UK}

% \author{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
% \author{Greg Chance}
% \author{Kevin McAreavey}
% \author{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
% \author{Tony Pipe}

\author{Greg Chance, Abanoub Ghobrial, Kevin McAreavey, S\'everin Lemaignan, Tony Pipe, Kerstin Eder 
\thanks{{\footnotesize
Manuscript  
received ...;
revised ...;  
accepted.... 
Date of publication ...;
date of current version ....

\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
%

This research has in part been funded by the ROBOPILOT and CAPRI projects. Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
% The Associate Editor for this paper was ....

Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
and 
Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.

S\'everin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
and
Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.

Digital Object Identifier ....
}}}
% \affil{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
% \affil{University of West of England, Bristol, UK}
% \affil{Bristol Robotics Laboratory, Bristol, UK}
%
\markboth{IEEE TRANSACTIONS ON INTELLIGENT TRANSPORTATION SYSTEMS VOL. ..., NO. ..., date...}{Chance, Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
%
\maketitle
%
\begin{abstract}
\noindent 
Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
%
A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history. 
%
Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
%
However, game engines are seldom deterministic. 
%
This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
%
A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results. 
%
Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work. 
%
Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.

% {\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}

% =========================================
% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
% %
% A key requirement for simulation-based development and verification is determinism.
% %
% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
% %
% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
% %
% However, game engines are seldom deterministic.

% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
% %
% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
% %
% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e. variance between repeats becomes negligible for development and testing work.

% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.

% =========================================
% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
% %
% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
% %
% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
% %
% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
% %
% However, game engines do not need to be deterministic per se. % when used for games. 
% %
% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively. 
% %
% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
% %
% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification. The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
% %
% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}

\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance. }
\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance. }
\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}

% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation. It is important for these engines to be deterministic in order for one to carry out these different processes. However, such engines are inherently non-deterministic. We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic. We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.          
\end{abstract}

\begin{IEEEkeywords}
Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
\end{IEEEkeywords}
\IEEEpeerreviewmaketitle

% ***************************************************
%  INTRODUCTION
% ***************************************************
\section{Introduction} \label{s:introduction}
% What is verification
% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
%
% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
% , offering a relatively inexpensive means to explore test cases while also
% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
%
Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.

Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}. Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}. In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.

% The use of game engines for AV testing
Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
%
CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.

% Requirements of simulation based verification SBV
State-of-the-art game engines provide a convenient option for simulation-based testing. They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses. 
%
Furthermore, they are easy to set up and run compared to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}. 
%
%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
% 
% GC Removing the following para:
%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models. CARLA currently offers various basic types of cameras and %ray-case based 
% lidar; extensions are still under development. 
% \todo{GC: what is ray-case? KIE: Not sure. :(}
%
Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
%
% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
%
% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. 
%
However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.

% \subsection{Contribution}
% In pursuit of an answer to this question we
In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
%
In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
% KIE: This is covered in the next sentence:
% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
% By analysing actor position variance 
We find that the CARLA simulator is non-deterministic. Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm. We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment. 
% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment. 
%
\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have. In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.  }
%
Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
%impact that using a game engine for AV simulation may have.
% on verification tasks. 
In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
%
\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have. In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this. }
%
% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
%We present a case study to illustrate the use of this method. The study highlights the effects of non-determinism of a game engine on simulation results. Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
%
%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
%

This paper is structured as follows.
%
Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
%
Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
%
Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
%
We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}

\todo[inline,color=green!40]{GC New below}

\section{Preliminaries} \label{s:prelim}

\subsection{Definitions}

Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
\\

\subsubsection{Determinism}
% What is a deterministic system
Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}. However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors. Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
% 
% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

A simulation can be thought of as the process of generating or producing experimental data. 
%
In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. 
%
If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e. all future states are perfectly reproducible from prior states and actions. 
%
However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data. \\

\begin{figure}[!t]
    \centering
    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
        \caption{Non-deterministic}
        \label{variance_description_a}
    \end{subfigure}

    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
        \caption{Deterministic}
        \label{variance_description_b}
    \end{subfigure}
    \caption{Demonstration of variance, precision, tolerance and determinism}
    \label{variance_description}
\end{figure}

% \subsection{Definitions: Precision, Variance \& Tolerance}
\subsubsection{Variance, Precision \& Tolerance}

We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
%
% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
% %
% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
%
% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value. \textit{Precision} is synonymous with \textit{variance} although inversely related mathematically. 
%
Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
%
\textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.

% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}. %
%
% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be? % when comparing the differences between multiple products. 
As an analogy, the simulator can be thought of as a manufacturing process that produces data. To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated. 
%
Those differences describe the spread or variance in the process output. A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
%
% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from. 
%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
%
Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.

If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design. If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
%

\subsubsection{Accuracy}
Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value. Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation. 
%
For a driving simulation, the reference value 
% , for which accuracy can be measured against, 
may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
%
In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica. In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}. \\

\subsubsection{Simulation Trace}
A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals. This definition could be extended to include other variables. %, but position is sufficient for this analysis.\\
A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\

\subsubsection{Simulation Variance \& Deviation}
If the simulator is non-deterministic then how can the simulation variance be measured? This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run. 
% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
For example, actor position variance is a distance-based metric that can be derived from simulation traces. The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained. 
% 
Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions. 
% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
%
%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
%
Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
%
% more intuitive to comprehend when considering verification tests, e.g. vehicle close-passing. Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
\\

\subsubsection{Scene, Scenario \& Situation}
We adopt the terminology defined for automated driving in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements. Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV, or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals. The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters. A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.

% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}

% Verification and Determinism 
% Why is determinism important for verification
%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
% KE Text
\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable. Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.  }
%

\subsection{When is Determinism needed?}
Determinism is a key requirement for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
%
Thus, a deterministic simulator has zero \textit{variance}.

A simulator with non-zero variance can no longer be considered deterministic. 
%
Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}. 
%
Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations. When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. 
%
This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.

% GC Text
\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing. A deterministic simulation environment guarantees that tests are repeatable. In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}. If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}. Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs. When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}

% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world. The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome. \textit{Tolerance} is the acceptable degree of variability between repeated execution outputs. When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy. }
%
%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution. Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour. 
%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
%

If the simulation is non-deterministic, e.g.\ 
% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
%e.g. it has a non-permissible variance in,
it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable. In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software. 
%

% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs. In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}

% Challenges of SBV
When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs % related to non-determinism 
a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions are:
%
\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
%
{\em How can one assess whether a simulation environment is deterministic?} and 
{\em How can one determine and control the simulation variance?}
\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}

% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject. Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.

% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.

% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\

% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.

% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\

% ======================
% This is all from "FrameworkAndChallenges" without using quotes 
% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system. Similarly, the use of simulation for traffic modelling is well established. Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\

% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require. On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure. 
% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.  
% ======================

% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc. are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}. \\\\
% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers. This previously was not important since in gaming, determinism is not critical. 
% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\

% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
% \begin{itemize}[leftmargin=*]
%     \item Necessary background on game engines used in AV simulations.
%     \item Review of sources of non-determinism in such engines.
%     \item Provide a method of how the level of non-determinism can be investigated and defined.
%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
% \end{itemize}

% ***************************************************
%  BACKGROUND
% ***************************************************
\section{Background} \label{s:background}

There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}. Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.

Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.  %However, when these engines are used for AV development then deterministic behaviour is required. 
%
Ensuring software operates deterministically is a non-trivial task.  Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
%
\todo[inline]{GC line added to introduction. We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}. Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion. But it is important that we demonstrate knowledge of others working on this in other areas. {\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
%
\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}. }
%
This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

% ======= Overview of a Game Loop
% \subsection{Game Engine Overview} \label{GameLoopSection}

>>>>

Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history. Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. However, game engines are seldom deterministic. This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results. Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work. Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.
